From Reversible Programs to \\ Univalent Universes and Back

### 2021-05-02 Pi+/paper/univalent.tex
More text in section 3
Fixes #73

We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts.

@@

@@

@@

\begin{definition}[Type families are fibrations]
  A type family $P : A \to \UU$ gives a fibration $\fst : \dsum{x:A}{P(x)} \to A$ with total space $\dsum{x:A}{P(x)}$
  and base space $A$ with a lifting property.
\end{definition}

\begin{definition}[$\term{transport-equiv}$]
  Given a type family $P : A \to \UU$, we define
  \[
    \term{transport-equiv} : x \id_{A} y \to P(x) \eqv P(y)
  \]
\end{definition}

\begin{definition}[Univalent Fibration]
  $P$ is a univalent fibration if $\term{transport-equiv}$ is an equivalence.
\end{definition}

>>>>

We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

@@

In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.

Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
by

\[
  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
\]

Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have

\[
  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\]

From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection $\pi_1 : \dsum{x:A}{P(x)} \to A$ to
the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have

\[
  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
\]

where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.

Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
$\term{transport-equiv}$.

\[
  \term{transport-equiv}(P) : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
\]

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or simply a univalent fibration) if
  $\term{transport-equiv}$ is an equivalence.
\end{definition}

@@

The univalence principle characterises paths in the universe. It says that equivalent types are equal, or the following
function is an equivalence.

\[
  \ua : A \id_{\UU} B \to A \eqv B
\]

Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.

@@
